---
title: "Gapminder Analysis"
author: "Maureen Rocas"
date: "Wednesday, February 04, 2015"
output: html_document
Source: Gapminder Data at http://www.gapminder.org/data/
---

This is a study of the Gapminder dataset for Unemployment Rate for Aged 25-54.
```{r, echo=FALSE}
#Initialize environment
#install.packages("reshape")

setwd("C:/Users/mrocas/Documents/docs/tutorials/udacity/Data Analysis with R/Lesson 3")
getwd()
library("ggplot2")
library("reshape")
library("gridExtra")
library("scales")

```

Load Gapminder data

```{r} 

gapminder <- read.csv('indicator_t 25-54 unemploy.csv')
names(gapminder)
dim(gapminder)

```

Let's clean up data by:
  - renaming 'Total.25.54.unemployment...." column to Country
  - removing the last column "X" as this contains NA value
  - removing the last row as this contain NULL

```{r}

gapminder$x
gapminder$Total.25.54.unemployment....

# rename column
gapminder <- rename(gapminder, c(Total.25.54.unemployment....="Country"))

#remove last column
gapminder <- subset(gapminder,select=-X)

#remove last row
gapminder <- gapminder[1:(nrow(gapminder)-1),]


#Add country_id
gapminder$CountryId <- c(1:29)

dim(gapminder)


```

Plot histogram

```{r}

#melt data by country
gapminder1 <- melt(gapminder[,1:27], id=c("Country", "CountryId"))

#Add a new Date Column, convert Year(variable) to a date dataype
gapminder1 <- transform(gapminder1, Date=rep(as.Date(variable,"X%Y")), nrow(gapminder1))

#remove rows with NA values
gapminder1NoNA <- subset(gapminder1, !is.na(value))

#Histogram
qplot(data=gapminder1NoNA,x=value, binwidth=0.50,
      xlab="Unemployment Rate (%)",
      ylab="Frequency in Sample Data",
      color=I("black"), fill=I("#099DD9")) + 
      scale_x_continuous(lim=c(0,22), breaks=seq(0,22,3))

```

The histogram shows that the concentration of data is around 3.5 to 6.5 unemployment rate.  Let's see the total number of observations per country.

```{r}

# Total Observations per Country
qplot(data=gapminder1NoNA,x=Country,
      xlab="Countries",
      main="Total Observations per Country",
      fill=ifelse(as.numeric(format(Date, "%Y")) < 1990, "1980s",ifelse(as.numeric(format(Date, "%Y")) < 2000, "1990s", "2000s")),
      geom="bar") + 
  ylab("Number of Observations") +
  scale_fill_manual(name="Date Range", values=c("1980s"="red","1990s"="blue","2000s"="green")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

```

As shown above, we don't have equal observations for each country.  Like Czech Republic and United Kingdome, there are missing observations for 1980s and 1990s.
We can also see that some countries have less data available for 1980s, like Switzerland, Australia, Germany, and others.

Here is a boxplot, showing the median and quantile values for each Country

```{r, fig.width=8, fig.height=6}

qplot(data=gapminder1NoNA, x=Country, y=value,
      geom="boxplot") +
  ylab("Unemployment Rate (%)") +
  coord_cartesian(ylim=c(0,22))  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

#Overall summary statistics
by(gapminder1NoNA$value, gapminder1NoNA$Country, summary)


```

Let's look at the Unemployment Rate per year per Country.  

Here we can see the decreasing or increasing unemployment rate per country per year
```{r, fig.width=8, fig.height=8}
#Unemployment Rate per Date
qplot(x=Date, y=Country,
      data=gapminder1NoNA, binwidth=1,
      xlab="Year", 
      size=value,
      geom="point",
      main="Unemployment Rate per Year per Country",
      ylab="Countries") +
  scale_size_continuous(name="Unemployment\nRate", breaks=seq(0,22, 4)) +
  scale_x_date(lim=c(as.Date("1981", "%Y"), as.Date("2005", "%Y")), 
                     breaks=date_breaks("2 years"),
                     labels=date_format("%Y"))

```
